---
title: "Extracting data from LTMP model"
author: "Denisse Fierro Arcos"
date: "2024-01-19"
output: 
  github_document:
    toc: true
    html_preview: true
---

# Goal of this notebook

This notebook will demonstrate how to access the Long-Term Monitoring Program (LTMP) dataset from AIMS. The goal of the LTMP is to measure the long-term status and trend of reefs in the Great Barrier Reef (GBR) World Heritage Area. Data has been collected across the GBR for over 35 years. There are a variety of variables measured during this campaign, but they all provide information about the health of the coral reefs.  
  
You can see more information about the LTMP in our [catalogue](https://stac.reefdata.io/browser/collections/aims-ltmp-mmp-coralreef/items/aims-ltmp-mmp-coralreef-model).  

  
# Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
#Accessing S3 bucket
library(arrow)
#Data manipulation
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
#Managing dates
library(lubridate)
#Plotting
library(ggplot2)
```

# Connecting to RIMReP collection
We will connect to the LTMP dataset from AIMS stored in a S3 bucket. Depending on your internet connection, this make take a couple of minutes.  
  
```{r connect}
#Establishing connection
data_bucket <- s3_bucket("s3://gbr-dms-data-public/aims-ltmp-mmp-coralreef-model/data.parquet")

#Accessing dataset
data_df <- open_dataset(data_bucket)
```
  
## Exploring dataset structure
We can explore the dataset `schema` to understand its structure, and identify the variables that are relevant to us.

```{r explore}
#Checking data structure
data_df$schema
```
    
We will get the first rows of the dataset to understand the information included in it because the column names are not very informative.  
  
```{r}
df_sub <- data_df |> 
  #We will ignore the geometry column for now
  select(!geometry) |> 
  #Get the first 10 rows
  slice_head(n = 10) |> 
  collect()

#Check results
names(df_sub)
```
  
We can see that there are a number of columns that are completely empty. We can remove these columns to make the dataset more manageable.  
  
```{r}
df_sub <- df_sub |> 
  #Remove columns with no data
  remove_empty("cols")

#Check results
df_sub
```
  
We can see that the dataset contains information about the date of the record, the location of the record (`domain_name` and `shelf`), the depth, the variable measured, among other things.

For this example, we will extract data for the year 2020, and we will calculate the mean abundance of fishes (`variable` columns) for each `domain_name`, `depth` and `shelf` combination.  
  
```{r}
data_mth <- data_df |> 
  #We will ignore the geometry column for now
  select(!geometry) |> 
  #Select data for 2020
  filter(report_year == 2020) |> 
  #Return results
  collect() |> 
  #Add a month column
  mutate(month = month(ymd(date), label = T)) |>
  remove_empty("cols") |> 
  #Group data
  group_by(month, depth, shelf, variable) |>
  #Calculate mean abundance
  summarise(mean_val = mean(mean, na.rm = T)) |> 
  ungroup()

#Check results
head(data_mth)
```

  
Let's check what the mean abundance of fishes looks like for each month.      
```{r}
data_mth |> 
  ggplot(aes(month, mean_val, color = variable))+
  geom_point()+
  facet_wrap(shelf~depth)
```
  
It appears that most of the data in 2020 came from offshore areas at a depth of 9 meters.  

By checking the contents of a subset of the dataset, we can get a better idea of the information included in it. This will help us to identify the variables that are relevant to us, and to understand the structure of the dataset.
  
## Extra example during workshop
Plotting hard coral cover over entire period of dataset.
  
```{r}
#Extracting hard coral data
hc <- data_df |>
  #Ignoring geometries (because it takes a long time)
  select(!geometry) |>
  filter(variable == "HARD CORAL") |> 
  collect()

#Creating a plot
hc |> 
  #Transforming to percentage
  mutate(mean = mean*100, lower = lower*100, upper = upper*100) |> 
  remove_empty("cols") |> 
  filter(domain_name == "LANGFORD AND BIRD ISLES" &
           domain_category == "reef" & purpose == "COMPOSITION") |> 
  mutate(date = ymd(date)) |> 
  ggplot(aes(date, mean, color = reefpage_category))+
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper, 
                  fill = reefpage_category), alpha = 0.3)+
  facet_wrap(~reefpage_category)+
  labs(fill = "Hard coral taxa", color = "Hard coral taxa",
       y = "Coral coverage (%)")+
  theme_classic()+
  theme(axis.title.x = element_blank(), legend.position = "top")
```
