---
title: "Extracting gridded data via the API"
author: "Denisse Fierro Arcos"
date: "2024-05-20"
output: 
  github_document:
    toc: true
    html_preview: true
---

# Extracting sea surface temperature from NOAA Coral Reef Watch

The [NOAA Coral Reef Watch Sea Surface Temperature (CRW SST)](https://coralreefwatch.noaa.gov/) is a high resolution (5 km) dataset that provides daily SST data for coral reef monitoring. This dataset is available in the DMS under the [**NOAA Coral Reef Watch** collection](https://stac.reefdata.io/browser/collections/noaa-crw). In this notebook, we will use access this dataset using the API. We will also show how to extract data for a specific area using a bounding box derived from a shapefile. For this step, we will use one of the geometries available in the GBR Complete Features shapefile, which is also available in the [DMS](https://stac.reefdata.io/browser/collections/gbrmpa-admin-regions/items/gbrmpa-complete-gbr-features). 
  
## Loading libraries
  
```{r, message = F, warning = F, message = F}
#Loading useful_functions script
source("useful_functions.R")
#Gridded data
library(terra)
library(tidyterra)
#Shapefiles
library(sf)
#Base maps
library(rnaturalearth)
#Data manupulation
library(tibble)
library(tidyr)
```
  
## Loading Whitsunday features from GBR shapefile
We will use the GBR features here, but we will only load boundaries for the Whitsunday for this exercise. However, the function loading these GBR features can load the entire shapefile or any other GBR feature.  
  
```{r}
#Loading boundaries for Whitsundays
whitsundays <- gbr_features(site_name = "Whitsunday")

#Checking results of query
whitsundays
```
  
## Connecting to RIMREP collection via API

From the STAC catalogue item for the [NOAA CRW SST](https://stac.reefdata.io/browser/collections/noaa-crw/items/noaa-crw-chs-sst?.language=en-AU), we can get the link to the API from the *Additional Resources* section of the page on the left under the map.  
  
**Note:** Before running the code chunk below, make sure you either have store your user credentials as environmental variables, or have this information with you to input in the `connect_dms_dataset` function below. Alternatively, if you already have an access token, you can provide this as an input in the `connect_dms_dataset` function. Refer to **The data API** subsection under **How to use DMS services and data** in the [README page](https://github.com/aodn/rimrep-training/blob/main/CoTS-training-Jan2024/README.md) for more information.  
  
If you do not user credentials, you will not be able to access our API, please contact the DMS team to set up an account by emailing [info-dms@utas.edu.au](mailto:info-dms@utas.edu.au).  
    
```{r}
#Defining API URL (obtained from STAC catalogue)
base_url <- "https://pygeoapi.reefdata.io/collections/noaa-crw-chs-sst/"

#Defining variable of interest (obtained from STAC catalogue)
variable_name <- "analysed_sst"

#Connecting to DMS and downloading data
sst_gbr <- connect_dms_dataset(base_url, variable_name, 
                               start_time = '1994-01-01', 
                               end_time = '2023-12-31',
                               bounding_shape = whitsundays)
```
  
## Plotting data 
We will plot the first layer of this raster in a map to check the temperature data.  
  
```{r}
#Get map of Australia
aust <- ne_countries(country = "Australia", returnclass = "sf")

#Start a plot
ggplot()+
  #Plot one raster layer
  geom_spatraster(data = sst_gbr$analysed_sst_1)+
  #Choose a nicer palette for our map
  scale_fill_distiller(palette = "YlOrRd", name = "SST")+
  #Add Australia
  geom_sf(data = aust)+
  #Add Whitsundays
  geom_sf(data = whitsundays, fill = NA, colour = "black")+
  #Establish map limits
  lims(x = c(148, 150), y = c(-20.5, -20))+
  #Apply a nice predefined theme
  theme_bw()+
  #Add a title
  labs(title = "Sea Surface Temperature")+
  #Center the plot title
  theme(plot.title = element_text(hjust = 0.5))
```
  
The grey areas are land areas without any temperature data.  

## Calculating monthly sea surface temperature mean
We will calculate the monthly mean of the sea surface temperature for the reefs surrounding the Whitsunday Island. We will use the `raster_calc` function included in the `useful_functions.R` script. The result will be gridded data.  
  
```{r}
#Applying function to calculate monthly mean
sst_mean <- raster_calc(sst_gbr, "monthly", "mean", na.rm = T)

#Checking result
sst_mean
```
  
We can see that the number of layers in the raster was reduced from `r nlyr(sst_gbr)` in the original raster downloaded from the API to `r nlyr(sst_mean)` when the monthly means were calculated.  

We can now plot the monthly mean for the sea surface temperature. We will do this for the first layer.  
  
```{r}
#Start a plot
ggplot()+
  #Plot one raster layer
  geom_spatraster(data = sst_mean$`mean_2019-01`)+
  #Choose a nicer palette for our map
  scale_fill_distiller(palette = "YlOrRd", name = "mean SST")+
  #Add Australia
  geom_sf(data = aust)+
  #Add Whitsundays
  geom_sf(data = whitsundays, fill = NA, colour = "black")+
  #Establish map limits
  lims(x = c(148, 150), y = c(-20.5, -20))+
  #Apply a nice predefined theme
  theme_bw()+
  #Add a title
  labs(title = "Mean SST")+
  #Center the plot title
  theme(plot.title = element_text(hjust = 0.5))
```
    
## Calculating time series and maximum monthly values
We can calculate the mean for each time step included in the raster to obtain a time series (i.e., a single value for each layer). Then we can identify the maximum value per month. The `ras_to_ts` function in the `useful_functions.R` script will do this for us. This function returns a list that includes the time series and the maximum monthly values from the time series.   
  
```{r}
#Calculating time series and maximum monthly values
sst_ts <- ras_to_ts(sst_mean, mean, na.rm = T) 

#Plotting time series
sst_ts$time_series |> 
  ggplot(aes(x = time, y = mean))+
  geom_line()+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y-%m",
               date_minor_breaks = "6 month")+
  theme_bw()+
  labs(title = "Monthly SST mean over Whitsunday Island",
       y = "Sea Surface Temperature (°C)")+
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_text(angle = 45, hjust = 1))
```
    
We can also plot the maximum monthly values. This function could be used to calculate the Maximum Monthly Mean (MMM), which is a metric used to assess the bleaching potential of the sea surface temperature. MMM is often calculated as the maximum monthly mean of the sea surface temperature over a 30-year period, but here we are showing how to apply the function with a smaller range of time.
  
```{r}
sst_ts$max_monthly_ts |> 
  #Note you need to add group = 1 to plot correctly
  ggplot(aes(month, max_monthly_val, group = 1))+
  geom_line()+
  theme_bw()+
  labs(title = "MMM for Whitsunday Island",
       y = "Sea Surface Temperature (°C)")+
  theme(axis.title.x = element_blank())
```
  
Plotting time series if months are numbers instead of names.  
  
```{r}
sst_ts$max_monthly_ts |> 
  #Adding column with numbers
  mutate(mth = 1:12) |> 
  ggplot(aes(mth, max_monthly_val))+
  geom_line()+
  #Add labels with abbreviated month names
  scale_x_continuous(limits = c(1, 12), breaks = 1:12,
                     labels = month.abb)+
  theme_bw()+
  labs(title = "MMM for Whitsunday Island",
       y = "Sea Surface Temperature (°C)")+
  theme(axis.title.x = element_blank())
```
  
# Extra examples during workshop
Extra code created during training.

Extracting data for a single point from Whitsunday Island raster.

```{r}
#Extracting data for a specific point
pt <- terra::extract(sst_gbr, vect("POINT (149 -20.2)"))

#Format data frame better for plotting
pt <- pt |> 
  #Pivot the table
  pivot_longer(!ID, names_to = "time", values_to = "sst") |> 
  #Adding time column - Get time from original raster
  mutate(time = time(sst_gbr),
         #Get time as year month
         year_month = format(time, "%Y-%m"))

#Calculating monthly means
pt_mean_sst <- pt |> 
  group_by(year_month) |>
  summarise(mean_sst = mean(sst, na.rm = T))

#Calculating maximum monthly values over entire period
pt_max_mth_sst <- pt_mean_sst |> 
  #Extracting month from year-month
  mutate(month = str_extract(year_month, "([0-9]{2})$",
                             group = 1)) |> 
  group_by(month) |> 
  summarise(monthly_max = max(mean_sst, na.rm = T))
  
#Calculating maximum SST by month over entire period
pt_max_mth_sst |> 
  ggplot(aes(x = month, y = monthly_max, group = 1))+
  geom_line()

#Plotting maximum monthly SST values
pt |> 
  group_by(year_month) |>
  summarise(max_sst_yr_mth = max(sst, na.rm = T)) |> 
  ggplot(aes(x = year_month, y = max_sst_yr_mth, group = 1))+
  geom_line()
```


  