---
title: "Extracting temperature using GBR features"
author: "Denisse Fierro Arcos"
date: "2023-06-14"
output: 
  md_document:
    variant: gfm
    toc: true
    preserve_yaml: true    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal of this notebook

This notebook will demonstrate how to use one or more Great Barrier Reef (GBR) features recognised by the Great Barrier Reef Marine Park Authority (GBRMPA) dataset to extract water temperature data from the AIMS Sea Surface Temperature Monitoring Program available at RIMReP `geoparquet` collection.

In addition to published `R` libraries, this notebook uses a set of functions especially created for manipulating RIMReP data collections. These functions are available in the `useful_functions.R` script, which **must** exist in the same directory where this notebook is located.

# Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
#Accessing S3 bucket
library(arrow)
#Data manipulation
library(dplyr)
library(magrittr)
library(stringr)
#Managing dates
library(lubridate)
#Plotting
library(ggplot2)
library(leaflet)
#Mapping
library(sf)
#Custom library
source("useful_functions.R")
```

# Connecting to RIMReP collection

Connecting to the AIMS Sea Surface Temperature Monitoring Program dataset. This can take a minute or so.

```{r connect}
#Establishing connection
data_bucket <- s3_bucket("s3://rimrep-data-public/091-aims-sst/test-50-64-spatialpart")

#Accessing dataset
data_df <- open_dataset(data_bucket)
```

# Extracting monitoring sites and their coordinates

```{r sites_bin}
sites_shp <- data_df %>% 
  #We select unique sites included in the dataset
  distinct(site, subsite, lon, lat) %>%
  #This will load them into memory
  collect()

#Creating shapefile of unique locations
sites_shp <- sites_shp %>% 
  #Creating column to identify deployment location
  mutate(deployment_location =  case_when(str_detect(subsite, "FL[0-9]{1}") ~ "reef flat",
                                          str_detect(subsite, "SL[0-9]{1}") ~ "reef slope",
                                          #If no condition is met, classify as "other"
                                          T ~ "other")) %>%
  #Turning into sf object
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

#Checking results
glimpse(sites_shp)
```
# Identifying AIMS sites within GBR feature of interest
In this section, we will use the functions included in the [useful_spatial_functions_data_extraction](https://github.com/aodn/rimrep-examples/blob/main/poc-data-api/useful_spatial_functions_data_extraction.R) script to load GBR features of interest. We will then use the boundaries of the feature of interest to identify the names of the sites within these limits.  
  
We will use the `gbr_features` function to access the boundaries for our sites of interest. If no parameters are given to this function, it will load all features recognised by GBRMPA. You could also provide the name of the site using the `site_name` parameter or the GBRMPA unique ID using the `site_id` parameter. You can also use both parameters at the same time in case you have the names for some sites and the IDs for others.  
  
As an example, we will use **U/N Reef** or unnamed reefs. It should be noted that **U/N Reef** refers to multiple unnamed reefs, and each reef can be identified by its `UNIQUE_ID`. We chose to use **U/N Reef** for this example simply because it will return multiple reefs within the GBR and thus will overlap with the most number of AIMS monitoring sites. 

```{r gbr_features}
un_reefs <- gbr_features(site_name = "u/n reef")

#Checking results
un_reefs
```
## Identifying sites of interest
Depending on the amount of polygons used to identify the sites of interest, this may take a couple of minutes.  
  
We will plot the original AIMS monitoring sites (black), the polygon of interest (pink) and the sites within polygons (blue) to check that sites have been correctly identified.

Note that unless the AIMS site is within a polygon, it will not to be identified in our final list.

Below we are including two ways of plotting sites and polygons: using `ggplot2`, which produces a static map, and using `leaflet`, which produces an interactive map.

### Mapping with `ggplot2`
```{r gg_map}
site_list <- sites_of_interest(sites_shp, un_reefs)

#Plotting results
#AIMS monitoring sites
sites_shp %>% 
  ggplot()+
  geom_sf()+
  #Polygon of interest
  geom_sf(inherit.aes = F, data = un_reefs, color = "pink")+
  #AIMS polygons within area of interest
  geom_sf(inherit.aes = F, data = site_list, color = "blue")+
  theme_bw()
```

### Mapping with `leaflet`

Maps produced with `leaflet` are interactive and allows users to zoom in and out as they need.

```{r leaflet_map, eval = F}
#Starting interactive map
leaflet() %>% 
  #Adding a basemap from Open Street Map
  addTiles() %>% 
  #Adding polygons of the unnamed (U/N) reefs
  addPolygons(data = un_reefs, weight = 1) %>% 
  #Adding all sites in black
  addCircleMarkers(data = sites_shp, radius = 1.5, color = "black") %>% 
  #Adding sites within U/N reefs in pink
  addCircleMarkers(data = site_list, radius = 3, color = "pink")
```

**Note**: Due to the size of the interactive map, we chose not to display it in the GitHub markdown (file ending in `.md`), however, you will be able to see it and interact with it when you run the code above in `RStudio` using the file ending in `.Rmd`.

## Extracting temperature for monitoring site within area of interest

We will now extract data only for the monitoring sites located within the GBR feature of our interest.

```{r sites_extract}
#Extracting data points within U/N reefs from the cloud
temp_area_int <- data_df %>%
  select(site, subsite, time, qc_val) %>% 
  inner_join(st_drop_geometry(site_list), by = c("site", "subsite")) %>% 
  collect()
```

Now we will calculate the monthly means for temperature. We will group the monthly means on deployment location. Note that points of interest within the U/N reefs were only deployed in reef flats, or reef slopes. There were no deployments in an unknown deployment location.

```{r calc_means}
temp_area_int <- temp_area_int %>% 
  #Adding year and month columns
  mutate(year = year(time),
         month = month(time)) %>%
  #Group by month, year and deployment location
  group_by(year, month, deployment_location) %>%
  #Calculating means for groups
  summarise(temp_monthly_mean = round(mean(qc_val, na.rm = TRUE), 2))

#Checking results
head(temp_area_int)
```

## Plotting timeseries

We can now use `ggplot2` to create a plot showing how temperature has changed over time. We will save the plot as a variable so we can save it to our local machine later.

```{r timeseries_plot}
temp_plot <- temp_area_int %>% 
  #Combining year and month columns into one
  mutate(date = ym(paste0(year, "-", month))) %>% 
  #Plotting temperature of y axis and time on x axis. Color data by site.
  ggplot(aes(x = date, y = temp_monthly_mean))+
  #Plot data as points and lines
  geom_point()+
  geom_line()+
  #Creating subplots for each site for easy comparison
  facet_grid(~deployment_location)+
  #Removing default grey background
  theme_bw()+
  #Change position of legend
  theme(legend.position = "top", axis.title.x = element_blank())+
  #Change position and title of legend
  # guides(colour = guide_legend(title.position = "top", title.hjust = 0.5, title = "Site names"))+
  labs(y = expression("Mean monthly temperature  " ( degree~C)))

#Checking plot
temp_plot
```

## Saving data summaries and plot
We will need to provide a path to the folder where we want to save our data summaries and plots, as well as the file names for our outputs.

```{r saving_outputs, eval = F}
#First we will provide a path to the folder where we want to save our data
folder_out <- "Outputs"

#Checking if folder exists, if not, create it
if(!dir.exists(folder_out)){
  dir.create(folder_out)
}

#Saving data summaries
data_out <- paste(folder_out, "monthly_means_temperature.csv", sep = "/")
write_csv_arrow(temp_area_int, file = data_out)

#Saving plot
plot_out <- paste(folder_out, "monthly_means_timeseries.png", sep = "/")
ggsave(plot_out, temp_plot, device = "png")
```

