---
title: "NOAA_Degree_Heating_Week"
author: "Denisse Fierro Arcos"
date: "2023-10-18"
output:
  github_document:
    toc: true
    html_preview: false
---

# Accessing and plotting NOAA's Degree Heating Week data
This notebook will demonstrate how to access the RIMReP collection for [NOAA's Coral Reef Watch - Degree Heating Week (DHW)](https://stac.staging.reefdata.io/browser/collections/noaa-crw/items/noaa-crw-dhw?.language=en&.asset=asset-data). This dataset provides coral bleaching heat stress index derived from satellite data a global scale with a temporal resolution of 1-day and a horizontal spatial resolution of about 5 km ($0.05^{\circ}$). The dataset available in the DMS is [version 3.1 of the NOAA's Coral Reef Watch - Degree Heating Week](https://coralreefwatch.noaa.gov/index.php), which includes data from January 1, 1985 to present.  
  
In this notebook, we will use the `connect_dms_dataset` function from the `useful_functions.R` script, which allows us to access data from the RIMReP DMS API. We also have the option of including spatial and temporal boundaries to extract the data of interest.  
    
# Loading libraries
  
```{r, results = "hide", message = F, warning = F}
#Loading useful_functions script
source("useful_functions.R")
#Mapping
library(terra)
```
  
# Connecting to RIMReP collection via API
From the STAC catalogue entry for the [DHW dataset](https://stac.staging.reefdata.io/browser/collections/noaa-crw/items/noaa-crw-dhw?.language=en&.asset=asset-data), we can get the link to the API from the *Additional Resources* section of the page (bottom left).  
  
We will show how to get data from this API link for the period between 2023-01-01 and 2023-01-07. For this example, we will select data for coastal waters up to 150 km away from the coastline between Daintree and Cairns.  
  
**Note:** Before running the code chunk below, ensure you head over to our dashboard: [https://dashboard.staging.reefdata.io/](https://dashboard.staging.reefdata.io/). Then click on the blue **Copy token to clipboard** button. After you run the code chunk below, you will see a pop-up box requesting for this token. Paste your token in the box and press *OK* to continue. If you do not have this token, you will not be able to access our API, please contact the RIMReP team to set up an account by emailing [info-rimrep@utas.edu.au](mailto:info-rimrep@utas.edu.au).  
  
```{r}
#Defining API URL (obtained from STAC catalogue)
base_url <- "https://pygeoapi.staging.reefdata.io/collections/noaa-crw-dhw/"

#Defining variable of interest (obtained from STAC catalogue)
variable_name <- "degree_heating_week"

#Connecting to DMS to extract data
ras_dhw <- connect_dms_dataset(base_url, variable_name,
                           start_time = "2023-01-01", end_time = "2023-01-07", 
                           lon_limits = c(145.30, 146.90),
                           lat_limits = c(-17, -16.30))
```
  
This function will query the API and return the data in gridded form (i.e., raster). Once the data is loaded to memory, we can plot it using the `terra` library.  
  
```{r}
plot(ras_dhw)
```


